``Interestingness'' in Mathematical Theory Exploration

### 2019-06-02 PhDThesis/benchmark.tex
Bunch more changes to thesis
Headingy things
Thesis updates

\section{Theory Exploration Benchmark}
\label{sec:proposal}

If we wish to make measurable progress on the task of conjecture generation, we
must quantify what it is we are trying to achieve. Various approaches can then
be measured and compared to optimise toward our goal. To this end we define a
benchmarking methodology, shown in Figure~\ref{figure:flow_chart} on
page~\pageref{figure:flow_chart}, which both generates a large
definition/ground-truth corpus, and provides a scalable, statistical approach to
evaluating MTE tools using this corpus. We follow a precision/recall approach
similar to prior work, with the main difference being the source of definitions
and ground truths: we take existing problem sets designed for automated theorem
proving, and adapt their content for use in the conjecture generation setting.

\subsection{Preparation}
\label{section:prep}

@@

\subsection{Sampling}
\label{section:sampling}

@@

\subsection{Evaluation}
\label{section:evaluation}

@@

  \end{tikzpicture}
  \caption[]{High-level view of the benchmarking methodology described in
    $\S$\ref{sec:proposal}, showing
    \begin{tikzpicture}
      \node [rectangle, text centered, draw=black]{processes};
    \end{tikzpicture},
    \begin{tikzpicture}
      \node [rectangle, rounded corners, text centered, draw=black]{data};
    \end{tikzpicture} and
    \begin{tikzpicture}
      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
             text centered, draw=black]{inputs};
  \end{tikzpicture}}
  \label{figure:flow_chart}
\end{figure}

@@

We have applied our benchmarking methodology to the QuickSpec and IsaCoSy MTE
tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem proving
benchmark as our ground truth corpus~\cite{claessen2015tip}.

To determine our benchmarking parameters we ran some initial tests on both tools
for a few samples sized between 1 and 100, for an hour each on our benchmarking
machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
of RAM. Most QuickSpec runs either finished within 200 seconds or not at all,
and sizes above 20 mostly timed out. IsaCoSy mostly finished within 300 seconds
on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
above this took thousands of seconds each, which we deemed infeasibly long.

Based on this we decided to benchmark sample sizes up to 20, since neither tool
seemed to perform well beyond that. The Speedup-Test protocol follows the
statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
so we pick 31 samples of each size in order to cross this threshold. This gave a
total of 1240 runs. To keep the benchmarking time down to a few days we chose a
timeout of 300 seconds, since that covered most of the successful QuickSpec and
IsaCoSy results we saw, and longer times gave rapidly diminishing
returns. During analysis, duplicate samples (caused by our requirement that
samples have a non-empty ground truth) were found and discarded, so only 14
samples of size 1 were used and 30 of size 2.

@@

TIP datatypes can have several ``constructors'' (introduction forms) and
``destructors'' (elimination forms; field accessors). For example the type of
lists from Figure~\ref{figure:list_theory} can be defined in the TIP format as
follows:

@@

TIP comes with software to translate its definitions into Haskell and Isabelle
code, including comparison functions and random data generators suitable for
QuickCheck. We translate all 269 unique definitions into a single module/theory
which is imported on each run of the tools, although only those functions which
appear in the current sample are included in the signature and explored. We also
encode all names in hexadecimal to avoid problems with language-specific naming
rules, for example \texttt{add} becomes \texttt{global616464} (the prefix
distinguishes these from local variables and prevents names from beginning with
a digit). This ensures that the generated conjectures will be using the same
names as the ground truth, rather than some language-specific variant.

\subsection{QuickSpec}

We benchmarked QuickSpec version 0.9.6, a tool written in Haskell for
conjecturing equations involving Haskell functions, described in more detail in
$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark QuickSpec, we
need to automate some of the decisions which are normally left up to the user:

\begin{sloppypar}
  \begin{itemize}
  \item We must decide what variables to include. We choose to add three
    variables for each type that appears as a function argument, except for
    types which have no QuickCheck data generators.
  \item We must \emph{monomorphise} all types. For example, functions like
    \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
    element type, but we need to pick a specific type in order to know which
    random value generator to use. We resolve this (arbitrarily) by picking
    \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
      \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
      \texttt{[]} (Haskell's list type constructor). If these violate some
      type class constraint, we pick a suitable type non-deterministically from
      those in scope during compilation; if no suitable type is found, we give
      up and don't include that function.}
  \item Haskell functions are ``black boxes'', which QuickSpec can't compare
    during its exploration process. They are also curried, always taking one
    argument but potentially returning another function. QuickSpec lets us
    assign an arity to each function in the signature, from 0 to 5, so we pick
    the highest that is type-correct, since this avoids a proliferation of
    incomparable, partially-applied functions.
  \end{itemize}
\end{sloppypar}

\begin{figure}
  \centering
  \input{quickspectime.pgf}
  \caption{Running times of QuickSpec on theories sampled from TIP. Each point
    is a successful run (spread out horizontally to reduce overlaps). Red
    crosses show runs which timed out, which occurred for every size. Each size
    has 31 runs total, except for size 1 (14 runs) and size 2 (30 runs) due to
    sampling restrictions. Box plots show inter-quartile range, which reaches
    the timeout for sizes above 8. Medians are marked with a line and remain
    near zero until size 10, with those of size 13 and 20 timing out. Whiskers
    show 1.5$\times$~inter-quartile range.}
  \label{figure:quickspec_runtimes}
\end{figure}

\begin{figure}
  \centering
  \input{quickspecprec.pgf}
  \caption{Precision and recall of successful QuickSpec runs (spread
    horizontally to reduce overlapping). Lines show the mean proportion for each
    sample size (average of the ratios) and errorbars show the sample standard
    deviation. Mean precision steadily reduces from around $0.5$ at size 1 to
    around $0.1$ for size 20, whilst mean recall remains roughly flat between
    $0.2$ and $0.5$"}
  \label{figure:quickspec_precRec}
\end{figure}

The time taken to explore samples of different sizes is shown in
Figure~\ref{figure:quickspec_runtimes}. Failed runs are shown in red: all
failures were due to timing out, and these occurred for all sample sizes but are
more common for larger samples (over half the runs for sample sizes 13 and 20
timed out). Runs which succeeded mostly finished within 30 seconds, generating
few or no conjectures; those taking longer generated more conjectures, with the
most being 133 conjectures from a sample of size 19.

Precision and recall are shown in Figure~\ref{figure:quickspec_precRec}.
Since QuickSpec generates monotonically more conjectures as definitions are
added to a signature (assuming sufficient testing), its decreasing precision
can't be due to finding fewer wanted conjectures at larger sizes. This is
supported by the relative flatness of the recall results. Rather, the number of
conjectures generated is increasing at a higher rate than the size of the ground
truth. These extra conjectures may involve those ``padding'' definitions in a
sample which don't contribute to its ground truth, or may be ``uninteresting''
relationships between the dependencies of different ground truth properties.

This indicates two potential improvements to the QuickSpec algorithm (as far as
this benchmark is concerned). The deluge of generated conjectures could be
filtered down to a more desirable sub-set by another post-processing step.
Alternatively, rather than exploring all of the given definitions together,
multiple smaller signatures could be selected from the input by predicting which
combinations are likely to lead to interesting conjectures; this would avoid
both the ``padding'' and the cross-dependency relationships. Both of these
methods could improve the precision, although care would be needed to avoid a
large reduction in recall. The latter option could also improve the running
time, since (based on Figure~\ref{figure:quickspec_runtimes}) multiple smaller
signatures may be faster to explore than a single large one. Such improvements
would also need to be ``generic'', to avoid over-fitting to this particular
benchmark.

QuickSpec's recall is limited by two factors: the algorithm is unable to
synthesise some properties, such as conditional equations, inequalities, terms
larger than the search depth and those containing anonymous functions. The
congruence closure algorithm used as a post-processor may also be removing
``interesting'' results, for example if we found an ``uninteresting'' result
which is more general.

\subsection{IsaCoSy}

We took IsaCoSy from version 2015.0.3 of the IsaPlanner project, and ran it
with the 2015 version of Isabelle. The following issues had to be overcome to
make our benchmark applicable to IsaCoSy:

\begin{itemize}
\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
  encoded in Isabelle. We strip out this type and the functions which depend on
  it before translating. It still appears in samples and contributes to the
  ground truth, which penalises IsaCoSy for being unable to explore such
  definitions.
\item When using a type in an IsaCoSy signature, that type's constructors will
  automatically be included in the exploration. Since those constructors will
  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
  even those may not be present in the current sample) this will unfairly reduce
  the calculated precision. To avoid this, we add a post-processing step which
  replaces all occurrences of a constructor with the corresponding wrapper, then
  discards any conjectures which involve functions other than those in the
  current sample. This presumably results in more work for IsaCoSy, exploring
  constructors unnecessarily, but it at least does not bring down the quality
  measures.
\item Since Isabelle is designed for theorem proving rather than programming, it
  requires every definition to be accompanied by proofs of exhaustiveness and
  termination. These are difficult to generate automatically, and don't exist in
  the case of partial functions like destructor wrappers. Hence we use the
  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
  the \texttt{sorry} keyword.
\item Partial functions cause problems during exploration, since they can throw
  an ``undefined'' exception which causes IsaCoSy to abort. We avoid this by
  pre-populating IsaCoSy's constraint set with these undefined expressions
  (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
  IsaCoSy from ever generating them.
\end{itemize}

\begin{figure}
  \centering
  \input{isacosytime.pgf}
  \caption{Running times of IsaCoSy on theories sampled from TIP. Each point
    is a successful run (spread out horizontally to reduce overlaps). Red
    crosses are failed runs, caused by timeouts or out-of-memory. All runs of
    size 1 succeeded, whilst nothing succeeded above size 6. Each size has 31
    runs total, except for size 1 (14 runs) and size 2 (30 runs) due to sampling
    restrictions. Box plots show inter-quartile range, which grows with size,
    and lines mark the median time, which increases rapidly from around 100
    seconds at size 1 to timing out at size 4 and above. Whiskers show
    1.5$\times$~inter-quartile range.}
  \label{figure:isacosy_runtimes}
\end{figure}

\begin{figure}
  \centering
  \input{isacosyprec.pgf}
  \caption{Precision and recall of all successful IsaCoSy runs (spread
    horizontally to reduce overlapping; all runs above size 6 failed). Lines
    show the mean proportion for each sample size (average of the ratios), which
    trends downwards for precision and appears flat for recall, although there
    are too few datapoints to be definitive. Errorbars show the sample standard
    deviation.}
  \label{figure:isacosy_precRec}
\end{figure}

The most striking result is how rapidly IsaCoSy's running time, shown in
Figure~\ref{figure:isacosy_runtimes}, increases with sample size: all runs above
size 6 failed, with most timing out and a few aborting early due to running out
of memory. Like in our preliminary testing, this increase appears exponential,
so even large increases to the timeout would not produce many more successful
observations. A significant amount of IsaCoSy's time (around 50 seconds) was
spent loading the generated theory (containing all distinct datatypes and
functions from TIP); this overhead is unfortunate, but since it is constant
across all sample sizes it does not affect our conclusions.

With so few successful datapoints it is difficult to make strong claims about
the precision/recall quality of IsaCoSy's output, shown in
Figure~\ref{figure:isacosy_precRec}. It is clear from the quantity of non-zero
proportions that IsaCoSy is capable of discovering interesting conjectures, and
the graphs appear to follow the same shapes as those of QuickSpec (decreasing
precision, flat recall), although the error margins are too wide to be
definitive.

@@

We compare the running times of QuickSpec and IsaCoSy using our paired variant
of the Speedup-Test protocol, where each of our sample sizes is a separate
``benchmark''. We used version 2 of the Speedup-Test \texttt{R} implementation,
which we patched to use the \emph{paired} form of the (one-sided) Wilcoxon
signed-rank test~\cite{wilcoxon1945individual}. We use Pratt's
method~\cite{pratt1959remarks} to handle ties, which occur when both tools time
out.

For each sample size (``benchmark'') the Speedup-Test protocol compares the
distributions of each tool's times using a Kolmogorov-Smirnov test. If these are
significantly different (with $\alpha < 0.05$), then the signed-rank test is
only performed if more than 30 samples are available. This was the case for all
sample sizes except for 1 and 2 (due to the removed duplicates), and hence those
two speedups were not deemed statistically significant. For all other sample
sizes QuickSpec was found to be significantly faster with $\alpha = 0.05$,
putting the proportion of sizes with faster median times between 66.9\% and
98.2\% with 95\% confidence. The magnitude of each speedup (median IsaCoSy time
divided by median QuickSpec time) is given in table~\ref{table:speedups}. We
would predict the speedup to grow for larger sample sizes, but the increasing
proportion of timeouts causes our estimates to become more conservative: by size
20 most runs are timing out, and hence are tied.

\begin{table}
  \centering
  \begin{tabular}{ |r|l| }
    \hline
    \bfseries Sample Size & \bfseries Speedup
    \csvreader[]{speedups.csv}{}
    {\\\hline\csvcoli&\csvcolii} \\
    \hline
  \end{tabular}
  \caption{Speedup from using QuickSpec compared to IsaCoSy, i.e. the median
    time taken by IsaCoSy divided by that of QuickSpec. QuickSpec was found to
    be significantly faster ($\alpha = 0.05$) for all sizes, except 1 and 2 due
    to too few samples. These are conservative estimates, since the 300 second
    time limit acts to reduce the measured time difference (e.g. sizes 13 and
    20, where the median for both tools timed out).}
  \label{table:speedups}
\end{table}

We compare the recall proportions by applying McNemar's test to only those
samples where both tools succeeded. We pooled these together from all sample
sizes to produce table~\ref{table:recall}, and found that the recall of
QuickSpec ($37\%$) is significantly higher than IsaCoSy ($25\%$) with a p-value
of 0.0026.

\begin{table}
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
    & \bfseries Found by IsaCoSy & \bfseries Not found by IsaCoSy \\
    \hline
    \bfseries Found by QuickSpec     & 28 & 19 \\
    \hline
    \bfseries Not found by QuickSpec & 4  & 75 \\
    \hline
  \end{tabular}
  \caption{Contingency table for ground truth properties, pooled from those
    samples where both QuickSpec and IsaCoSy finished successfully. The combined
    recall for QuickSpec is $37\%$ and for IsaCoSy is $25\%$.}
  \label{table:recall}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
              & \bfseries Interesting & \bfseries Uninteresting \\
    \hline
    \bfseries IsaCoSy   & 32          & 137      \\
    \hline
    \bfseries QuickSpec & 47          & 301      \\
    \hline
  \end{tabular}
  \caption{Interesting and uninteresting conjectures generated by QuickSpec and
    IsaCoSy, pooled from those samples where both both finished successfully.
    The combined precision for QuickSpec is $14\%$ and for IsaCoSy is $19\%$.}
  \label{table:precision}
\end{table}

We tested these totals for independence using Boschloo's test and find a p-value
of 0.111, which exceeds our (conventional) significance threshold of
$\alpha = 0.05$; hence we do not consider the difference between these
proportions (14\% for QuickSpec, 19\% for IsaCoSy) to be significant.

Note that precision is the only comparison where IsaCoSy scored higher, and even
that was not found to be significant. Also, precision on its own is not the
whole story: it can be increased at the expense of recall by making a tool more
conservative. IsaCoSy may already be overly-conservative compared to QuickSpec,
given that its recall is significantly lower.

More importantly, the poor time and memory usage of IsaCoSy meant that very few
samples finished successfully. If we include failed runs in our tables, treating
them as if they succeeded with no output, the resulting statistics lean
overwhelmingly in favour of QuickSpec simply because it generated so much more
than IsaCoSy in the available time.

\section{Discussion}
\label{sec:discussion}

@@

Another practical limitation of our benchmarking approach is that it only
applies to tools which act in ``batch mode'', i.e. those which choose to halt
after emitting some output. Whilst all of the systems we have encountered are of
this form, some (such as IsaCoSy, QuickSpec 2 and Speculate) could conceivably
be run without a depth limit, and form part of the ``scientist assistant'' role
which Lenat envisaged for AM, or McCarthy's earlier ``advice
taker''~\cite{McCarthy_Programs59}. Analogous benchmarks could be designed for
such long-running, potentially interactive programs, but that is beyond the
scope of this project.

>>>>

\chapter{Theory Exploration Benchmark}
\label{sec:benchmark}

If we wish to make measurable progress on the task of conjecture generation, we
must quantify what it is we are trying to achieve. Various approaches can then
be measured and compared to optimise toward our goal. To this end we define a
benchmarking methodology, shown in Figure~\ref{fig:flow_chart} on
page~\pageref{fig:flow_chart}, which both generates a large
definition/ground-truth corpus, and provides a scalable, statistical approach to
evaluating MTE tools using this corpus. We follow a precision/recall approach
similar to prior work, with the main difference being the source of definitions
and ground truths: we take existing problem sets designed for automated theorem
proving, and adapt their content for use in the conjecture generation setting.

\section{Preparation}
\label{section:prep}

@@

\section{Sampling}
\label{section:sampling}

@@

\section{Evaluation}
\label{section:evaluation}

@@

  \end{tikzpicture}
  \caption[]{High-level view of the benchmarking methodology described in
    $\S$\ref{sec:benchmark}, showing
    \begin{tikzpicture}
      \node [rectangle, text centered, draw=black]{processes};
    \end{tikzpicture},
    \begin{tikzpicture}
      \node [rectangle, rounded corners, text centered, draw=black]{data};
    \end{tikzpicture} and
    \begin{tikzpicture}
      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
             text centered, draw=black]{inputs};
  \end{tikzpicture}}
  \label{fig:flow_chart}
\end{figure}

@@

We have applied our benchmarking methodology to the \quickspec{} and \isacosy{}
MTE tools, using version 0.2 of the TIP (Tons of Inductive Problems) theorem
proving benchmark as our ground truth corpus~\cite{claessen2015tip}.

To determine our benchmarking parameters we ran some initial tests on both tools
for a few samples sized between 1 and 100, for an hour each on our benchmarking
machine with a 3.2GHz dual-core Intel i5 processor with hyper-threading and 8GB
of RAM. Most \quickspec{} runs either finished within 200 seconds or not at all,
and sizes above 20 mostly timed out. \isacosy{} mostly finished within 300 seconds
on sample sizes up to 4, but by size 8 was mostly timing out; its few successes
above this took thousands of seconds each, which we deemed infeasibly long.

Based on this we decided to benchmark sample sizes up to 20, since neither tool
seemed to perform well beyond that. The Speedup-Test protocol follows the
statistical ``rule of thumb'' of treating sample sizes $\leq$ 30 as ``small'',
so we pick 31 samples of each size in order to cross this threshold. This gave a
total of 1240 runs. To keep the benchmarking time down to a few days we chose a
timeout of 300 seconds, since that covered most of the successful \quickspec{} and
\isacosy{} results we saw, and longer times gave rapidly diminishing
returns. During analysis, duplicate samples (caused by our requirement that
samples have a non-empty ground truth) were found and discarded, so only 14
samples of size 1 were used and 30 of size 2.

@@

TIP datatypes can have several ``constructors'' (introduction forms) and
``destructors'' (elimination forms; field accessors). For example the type of
lists from Figure~\ref{fig:list_theory} can be defined in the TIP format as
follows:

@@

TIP comes with software to translate its definitions into Haskell and Isabelle
code, including comparison functions and random data generators suitable for
\quickcheck{}. We translate all 269 unique definitions into a single module/theory
which is imported on each run of the tools, although only those functions which
appear in the current sample are included in the signature and explored. We also
encode all names in hexadecimal to avoid problems with language-specific naming
rules, for example \texttt{add} becomes \texttt{global616464} (the prefix
distinguishes these from local variables and prevents names from beginning with
a digit). This ensures that the generated conjectures will be using the same
names as the ground truth, rather than some language-specific variant.

\subsection{\quickspec{}}

We benchmarked \quickspec{} version 0.9.6, a tool written in Haskell for
conjecturing equations involving Haskell functions, described in more detail in
$\S$\ref{sec:existing-tools}. In order to thoroughly benchmark \quickspec{}, we
need to automate some of the decisions which are normally left up to the user:

\begin{sloppypar}
  \begin{itemize}
  \item We must decide what variables to include. We choose to add three
    variables for each type that appears as a function argument, except for
    types which have no \quickcheck{} data generators.
  \item We must \emph{monomorphise} all types. For example, functions like
    \texttt{constructor-Cons} are \emph{polymorphic}: they build lists of any
    element type, but we need to pick a specific type in order to know which
    random value generator to use. We resolve this (arbitrarily) by picking
    \texttt{Integer}.~\footnote{We pick \texttt{Integer} for variables of kind
      \texttt{*} (types); for kind \texttt{* -> *} (type constructors) we pick
      \texttt{[]} (Haskell's list type constructor). If these violate some
      type class constraint, we pick a suitable type non-deterministically from
      those in scope during compilation; if no suitable type is found, we give
      up and don't include that function.}
  \item Haskell functions are ``black boxes'', which \quickspec{} can't compare
    during its exploration process. They are also curried, always taking one
    argument but potentially returning another function. \quickspec{} lets us
    assign an arity to each function in the signature, from 0 to 5, so we pick
    the highest that is type-correct, since this avoids a proliferation of
    incomparable, partially-applied functions.
  \end{itemize}
\end{sloppypar}

\begin{figure}
  \centering
  \input{quickspectime.pgf}
  \caption{Running times of \quickspec{} on theories sampled from TIP. Each point
    is a successful run (spread out horizontally to reduce overlaps). Red
    crosses show runs which timed out, which occurred for every size. Each size
    has 31 runs total, except for size 1 (14 runs) and size 2 (30 runs) due to
    sampling restrictions. Box plots show inter-quartile range, which reaches
    the timeout for sizes above 8. Medians are marked with a line and remain
    near zero until size 10, with those of size 13 and 20 timing out. Whiskers
    show 1.5$\times$~inter-quartile range.}
  \label{fig:quickspec_runtimes}
\end{figure}

\begin{figure}
  \centering
  \input{quickspecprec.pgf}
  \caption{Precision and recall of successful \quickspec{} runs (spread
    horizontally to reduce overlapping). Lines show the mean proportion for each
    sample size (average of the ratios) and errorbars show the sample standard
    deviation. Mean precision steadily reduces from around $0.5$ at size 1 to
    around $0.1$ for size 20, whilst mean recall remains roughly flat between
    $0.2$ and $0.5$"}
  \label{fig:quickspec_precRec}
\end{figure}

The time taken to explore samples of different sizes is shown in
Figure~\ref{fig:quickspec_runtimes}. Failed runs are shown in red: all
failures were due to timing out, and these occurred for all sample sizes but are
more common for larger samples (over half the runs for sample sizes 13 and 20
timed out). Runs which succeeded mostly finished within 30 seconds, generating
few or no conjectures; those taking longer generated more conjectures, with the
most being 133 conjectures from a sample of size 19.

Precision and recall are shown in Figure~\ref{fig:quickspec_precRec}.
Since \quickspec{} generates monotonically more conjectures as definitions are
added to a signature (assuming sufficient testing), its decreasing precision
can't be due to finding fewer wanted conjectures at larger sizes. This is
supported by the relative flatness of the recall results. Rather, the number of
conjectures generated is increasing at a higher rate than the size of the ground
truth. These extra conjectures may involve those ``padding'' definitions in a
sample which don't contribute to its ground truth, or may be ``uninteresting''
relationships between the dependencies of different ground truth properties.

This indicates two potential improvements to the \quickspec{} algorithm (as far as
this benchmark is concerned). The deluge of generated conjectures could be
filtered down to a more desirable sub-set by another post-processing step.
Alternatively, rather than exploring all of the given definitions together,
multiple smaller signatures could be selected from the input by predicting which
combinations are likely to lead to interesting conjectures; this would avoid
both the ``padding'' and the cross-dependency relationships. Both of these
methods could improve the precision, although care would be needed to avoid a
large reduction in recall. The latter option could also improve the running
time, since (based on Figure~\ref{fig:quickspec_runtimes}) multiple smaller
signatures may be faster to explore than a single large one. Such improvements
would also need to be ``generic'', to avoid over-fitting to this particular
benchmark.

\quickspec{}'s recall is limited by two factors: the algorithm is unable to
synthesise some properties, such as conditional equations, inequalities, terms
larger than the search depth and those containing anonymous functions. The
congruence closure algorithm used as a post-processor may also be removing
``interesting'' results, for example if we found an ``uninteresting'' result
which is more general.

\subsection{\isacosy{}}

We took \isacosy{} from version 2015.0.3 of the IsaPlanner project, and ran it
with the 2015 version of Isabelle. The following issues had to be overcome to
make our benchmark applicable to \isacosy{}:

\begin{itemize}
\item TIP includes a benchmark called \texttt{polyrec} whose types cannot be
  encoded in Isabelle. We strip out this type and the functions which depend on
  it before translating. It still appears in samples and contributes to the
  ground truth, which penalises \isacosy{} for being unable to explore such
  definitions.
\item When using a type in an \isacosy{} signature, that type's constructors will
  automatically be included in the exploration. Since those constructors will
  not appear in the ground truth (we use $\eta$-expanded wrappers instead, and
  even those may not be present in the current sample) this will unfairly reduce
  the calculated precision. To avoid this, we add a post-processing step which
  replaces all occurrences of a constructor with the corresponding wrapper, then
  discards any conjectures which involve functions other than those in the
  current sample. This presumably results in more work for \isacosy{}, exploring
  constructors unnecessarily, but it at least does not bring down the quality
  measures.
\item Since Isabelle is designed for theorem proving rather than programming, it
  requires every definition to be accompanied by proofs of exhaustiveness and
  termination. These are difficult to generate automatically, and don't exist in
  the case of partial functions like destructor wrappers. Hence we use the
  ``quick and dirty'' option in Isabelle, which lets us skip these proofs with
  the \texttt{sorry} keyword.
\item Partial functions cause problems during exploration, since they can throw
  an ``undefined'' exception which causes \isacosy{} to abort. We avoid this by
  pre-populating \isacosy{}'s constraint set with these undefined expressions
  (for example \texttt{(destructor-head constructor-Nil)}), hence preventing
  \isacosy{} from ever generating them.
\end{itemize}

\begin{figure}
  \centering
  \input{isacosytime.pgf}
  \caption{Running times of \isacosy{} on theories sampled from TIP. Each point
    is a successful run (spread out horizontally to reduce overlaps). Red
    crosses are failed runs, caused by timeouts or out-of-memory. All runs of
    size 1 succeeded, whilst nothing succeeded above size 6. Each size has 31
    runs total, except for size 1 (14 runs) and size 2 (30 runs) due to sampling
    restrictions. Box plots show inter-quartile range, which grows with size,
    and lines mark the median time, which increases rapidly from around 100
    seconds at size 1 to timing out at size 4 and above. Whiskers show
    1.5$\times$~inter-quartile range.}
  \label{fig:isacosy_runtimes}
\end{figure}

\begin{figure}
  \centering
  \input{isacosyprec.pgf}
  \caption{Precision and recall of all successful \isacosy{} runs (spread
    horizontally to reduce overlapping; all runs above size 6 failed). Lines
    show the mean proportion for each sample size (average of the ratios), which
    trends downwards for precision and appears flat for recall, although there
    are too few datapoints to be definitive. Errorbars show the sample standard
    deviation.}
  \label{fig:isacosy_precRec}
\end{figure}

The most striking result is how rapidly \isacosy{}'s running time, shown in
Figure~\ref{fig:isacosy_runtimes}, increases with sample size: all runs above
size 6 failed, with most timing out and a few aborting early due to running out
of memory. Like in our preliminary testing, this increase appears exponential,
so even large increases to the timeout would not produce many more successful
observations. A significant amount of \isacosy{}'s time (around 50 seconds) was
spent loading the generated theory (containing all distinct datatypes and
functions from TIP); this overhead is unfortunate, but since it is constant
across all sample sizes it does not affect our conclusions.

With so few successful datapoints it is difficult to make strong claims about
the precision/recall quality of \isacosy{}'s output, shown in
Figure~\ref{fig:isacosy_precRec}. It is clear from the quantity of non-zero
proportions that \isacosy{} is capable of discovering interesting conjectures, and
the graphs appear to follow the same shapes as those of \quickspec{} (decreasing
precision, flat recall), although the error margins are too wide to be
definitive.

@@

We compare the running times of \quickspec{} and \isacosy{} using our paired variant
of the Speedup-Test protocol, where each of our sample sizes is a separate
``benchmark''. We used version 2 of the Speedup-Test \texttt{R} implementation,
which we patched to use the \emph{paired} form of the (one-sided) Wilcoxon
signed-rank test~\cite{wilcoxon1945individual}. We use Pratt's
method~\cite{pratt1959remarks} to handle ties, which occur when both tools time
out.

For each sample size (``benchmark'') the Speedup-Test protocol compares the
distributions of each tool's times using a Kolmogorov-Smirnov test. If these are
significantly different (with $\alpha < 0.05$), then the signed-rank test is
only performed if more than 30 samples are available. This was the case for all
sample sizes except for 1 and 2 (due to the removed duplicates), and hence those
two speedups were not deemed statistically significant. For all other sample
sizes \quickspec{} was found to be significantly faster with $\alpha = 0.05$,
putting the proportion of sizes with faster median times between 66.9\% and
98.2\% with 95\% confidence. The magnitude of each speedup (median \isacosy{} time
divided by median \quickspec{} time) is given in table~\ref{table:speedups}. We
would predict the speedup to grow for larger sample sizes, but the increasing
proportion of timeouts causes our estimates to become more conservative: by size
20 most runs are timing out, and hence are tied.

\begin{table}
  \centering
  \begin{tabular}{ |r|l| }
    \hline
    \bfseries Sample Size & \bfseries Speedup
    \csvreader[]{speedups.csv}{}
    {\\\hline\csvcoli&\csvcolii} \\
    \hline
  \end{tabular}
  \caption{Speedup from using \quickspec{} compared to \isacosy{}, i.e. the median
    time taken by \isacosy{} divided by that of \quickspec{}. \quickspec{} was found to
    be significantly faster ($\alpha = 0.05$) for all sizes, except 1 and 2 due
    to too few samples. These are conservative estimates, since the 300 second
    time limit acts to reduce the measured time difference (e.g. sizes 13 and
    20, where the median for both tools timed out).}
  \label{table:speedups}
\end{table}

We compare the recall proportions by applying McNemar's test to only those
samples where both tools succeeded. We pooled these together from all sample
sizes to produce table~\ref{table:recall}, and found that the recall of
\quickspec{} ($37\%$) is significantly higher than \isacosy{} ($25\%$) with a p-value
of 0.0026.

\begin{table}
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
    & \bfseries Found by \isacosy{} & \bfseries Not found by \isacosy{} \\
    \hline
    \bfseries Found by \quickspec{}     & 28 & 19 \\
    \hline
    \bfseries Not found by \quickspec{} & 4  & 75 \\
    \hline
  \end{tabular}
  \caption{Contingency table for ground truth properties, pooled from those
    samples where both \quickspec{} and \isacosy{} finished successfully. The combined
    recall for \quickspec{} is $37\%$ and for \isacosy{} is $25\%$.}
  \label{table:recall}
\end{table}

@@

\begin{table}
  \centering
  \begin{tabular}{ |r|l|l| }
    \hline
              & \bfseries Interesting & \bfseries Uninteresting \\
    \hline
    \bfseries \isacosy{}   & 32          & 137      \\
    \hline
    \bfseries \quickspec{} & 47          & 301      \\
    \hline
  \end{tabular}
  \caption{Interesting and uninteresting conjectures generated by \quickspec{} and
    \isacosy{}, pooled from those samples where both both finished successfully.
    The combined precision for \quickspec{} is $14\%$ and for \isacosy{} is $19\%$.}
  \label{table:precision}
\end{table}

We tested these totals for independence using Boschloo's test and find a p-value
of 0.111, which exceeds our (conventional) significance threshold of
$\alpha = 0.05$; hence we do not consider the difference between these
proportions (14\% for \quickspec{}, 19\% for \isacosy{}) to be significant.

Note that precision is the only comparison where \isacosy{} scored higher, and even
that was not found to be significant. Also, precision on its own is not the
whole story: it can be increased at the expense of recall by making a tool more
conservative. \isacosy{} may already be overly-conservative compared to \quickspec{},
given that its recall is significantly lower.

More importantly, the poor time and memory usage of \isacosy{} meant that very few
samples finished successfully. If we include failed runs in our tables, treating
them as if they succeeded with no output, the resulting statistics lean
overwhelmingly in favour of \quickspec{} simply because it generated so much more
than \isacosy{} in the available time.

\subsection{Discussion}
\label{sec:discussion}

@@

Another practical limitation of our benchmarking approach is that it only
applies to tools which act in ``batch mode'', i.e. those which choose to halt
after emitting some output. Whilst all of the systems we have encountered are of
this form, some (such as \isacosy{}, \quickspec{} 2 and \speculate{}) could
conceivably be run without a depth limit, and form part of the ``scientist
assistant'' role which Lenat envisaged for AM, or McCarthy's earlier ``advice
taker''~\cite{McCarthy_Programs59}. Analogous benchmarks could be designed for
such long-running, potentially interactive programs, but that is beyond the
scope of this project.
